---
title: "Cohen's d Calculator"
subtitle: "Calculating and Visualizing Effect Sizes."
author: "Andy Grogan-Kaylor" 
categories: [stats]
date: "2024-2-5"
---

# Effect Sizes 

Let's imagine that you are studying an outcome like a *mental health outcome*. 

Suppose that your control group has a mean score of 100, and a standard deviation of 5 on this outcome. 

Suppose that your treatment group has a mean score of 110, and a standard deviation of 7 on this outcome.

```{r}
#| echo: true
#| code-fold: true

options(scipen = 999) # turn off scientific notation

N <- 100 # sample size

control <- rnorm(N, 100, 5) # control group data

treatment <- rnorm(N, 110, 7) # control group data

```

```{r}
#| echo: false
#| fig-height: 3

library(ggplot2)

library(michigancolors)

x <- c(control, treatment)

group <- c(rep("control", N),
           rep("treatment", N))

mydata <- data.frame(x, group)

ggplot(data = mydata,
       aes(x = x,
           fill = group)) +
  geom_histogram(alpha = .5) +
  scale_fill_manual(values = c(michigancolors("blue"),
                               michigancolors("maize"))) +
  theme_minimal()
  
```

# Statistical Significance

It's relatively easy to test for the *statistical significance* of this difference, as one can see in the example below.

```{r}
#| echo: true
#| code-fold: true

t.test(treatment, control) # t-test of treatment vs. control

```

# Substantive Significance

But what if you want to think about the *substantive significance* of the difference between treatment and control group? Effect sizes are *one way* to think about these issues. Discussion about the *merits of*, and *calculation of* effect sizes is *energetic* and *complex*. However, one commonly accepted way of thinking about effect sizes is *Cohen's d*.

# Cohen's d Calculator

I have recently updated my calculator to calculate and visualize Cohen's d. It can be found here: [https://agrogan.shinyapps.io/es_calculator/](https://agrogan.shinyapps.io/es_calculator/).











