---
title: "World Bank Data Demo"
author: "Andy Grogan-Kaylor"
date: "2026-03-30"
categories: [stats, R]
code-fold: true
code-summary: "Show the code"
---

# Background

The [World Bank](http://www.worldbank.org/) collects statistical information from countries around the world.  A particularly useful data set is the **W**orld **D**evelopment **I**ndicators [(WDI)](http://data.worldbank.org/data-catalog/world-development-indicators) which are country level statistical information from around the world.

R is unique in that using `library(WDI)` you can download indicator data directly from the World Bank, read it into a data set, and put it to use.  Using `library(plotly)` you can even make cool looking motion charts, somewhat reminiscent of those popularized by Hans Rosling.

While the code below is seemingly arcane, it is important to recognize that it is simple in structure. It is very possible to re-purpose the code below using some of the many *1,000's* of WDI indicators that are of interest to you.

# Call Relevant Libraries

```{r}
#| output: false
#| echo: true

library(WDI) # for accessing World Bank data

library(dplyr) # data wrangling

library(ggplot2) # beautiful graphs

library(plotly) # cool Hans Rosling style visualizations

library(DT) # data table

options(scipen = 999) # penalize scientific notation

```

```{r}
#| output: false
#| echo: false

library(foreign) # exporting to Stata

library(haven) # exporting to Stata

```

# Get Some Data From the World Development Indicators (WDI)

```{r}
#| eval: false
#| echo: true

# get names of specific indicators from WDI Data Catalog

mydata <- WDI(country="all", 
              indicator=c("SI.POV.GINI", # Gini
                          "NY.GDP.PCAP.CD", # GDP
                          "SE.ADT.LITR.ZS", # adult literacy
                          "SP.DYN.LE00.IN", # life expectancy
                          "VC.BTL.DETH", # battle related deaths
                          "SP.POP.TOTL", # population
                          "SN.ITK.DEFC.ZS"), # undernourishment
              start = 1980, 
              end = 2023, 
              extra = TRUE) 


save(mydata, file="WorldBankData.RData")

```

# Rename Some Variables

```{r}
#| eval: false
#| echo: true

# think about renaming some variables with more intuitive names
# e.g....

# rename some variables with dplyr (just copy and paste your indicators)

mydata <- dplyr::rename(mydata, 
                        GDP = NY.GDP.PCAP.CD,
                        adult_literacy = SE.ADT.LITR.ZS,
                        life_expectancy = SP.DYN.LE00.IN, 
                        battle_death = VC.BTL.DETH,
                        population = SP.POP.TOTL,
                        Gini = SI.POV.GINI,
                        undernourishment = SN.ITK.DEFC.ZS)

mydata <- mydata %>% 
  filter(region != "Aggregates") # remove Aggregates

mydata$country_name <- mydata$country

mydata$country <-as.factor(mydata$country)

save(mydata, file="WorldBankData.RData")

```

```{r}
#| echo: false
#| eval: false

write_dta(mydata, "WorldBankData.dta")

# write.dta(mydata, "WorldBankData.dta")

write.csv(mydata, 
          file = "WorldBankData.csv",
          row.names = FALSE)

```

```{r}
#| echo: false

load("WorldBankData.RData")

```

# Take a Quick Look at the Data

```{r}
#| echo: true

# head(mydata) # look at the data

mydata %>% 
  select(country, 
         region,
         year,
         GDP, 
         adult_literacy,
         life_expectancy, 
         battle_death,
         population, 
         Gini,
         undernourishment) %>%
  datatable(rownames = FALSE,
            filter = 'top', 
            extensions = 'Buttons', 
            options = list(
              dom = 'Bfrtip',
              buttons = c('copy', 'csv', 'excel', 'pdf', 'print')),
              caption = 'World Bank Data')

```

# ggplot Graphs

```{r}
#| warning: false
#| echo: true

p1 <- ggplot(mydata,
       aes(x = GDP,
           y = life_expectancy,
           color = region)) +
  geom_point() +
  geom_smooth() +
  scale_color_viridis_d() +
  labs(title = "Life Expectancy by GDP",
       x = "GDP",
       y = "Life Expectancy")

p1

p1 + facet_wrap(~region)

```

# Map

```{r}
#| warning: false
#| echo: true

mymap <- mydata %>% 
  filter(year == 2020) %>% 
  plot_geo(locations = ~iso3c,
           color = ~life_expectancy,
           z = ~life_expectancy,
           text = ~country) %>% 
  layout(title = "Countries by Life Expectancy in 2020", 
         geo = list(showland = FALSE,
                    showcountries = TRUE)) %>%
  colorbar(title = 'life expectancy') 
         
mymap
         
```

# Globe

```{r}
#| warning: false
#| echo: true

myglobe <- mydata %>% 
  filter(year == 2020) %>% 
  plot_geo(locations = ~iso3c,
           color = ~life_expectancy,
           z = ~life_expectancy,
           text = ~country) %>% 
  layout(title = "Countries by Life Expectancy in 2020", 
         geo = list(showland = FALSE,
                    showcountries = TRUE,
                    projection = list(type = 'orthographic',
                                      rotation = list(lon = -30,
                                                      lat = 10,
                                                      roll = 0)))) %>%
  colorbar(title = 'life expectancy') 


myglobe

```

# Plot the Motion Charts

```{r}
#| echo: true

mydata <- mydata %>% 
  filter(region != "Aggregates") # remove aggregates

```

## Life Expectancy by Year

::: {.panel-tabset}

### `ggplot` with `ggplotly`

```{r}
#| echo: true

p0 <- ggplot(mydata,
       aes(x = year,
           y = life_expectancy,
           color = region,
           size = population,
           frame = year,
           label = country)) + 
  geom_point() +
  labs(title = "Life Expectancy by Year",
       x = "Year",
       y = "Life Expectancy") +
  scale_color_discrete(name = "Region")

ggplotly(p0)

```

### `plotly`

```{r}
#| warning: false
#| echo: true

p1 <- plot_ly(mydata,
              x = ~year, 
              y = ~life_expectancy, 
              size = ~population, 
              fill = ~'',
              color = ~region,
              frame = ~year, 
              text = ~country, 
              hoverinfo = "text",
              type = 'scatter',
              mode = 'markers',
              showlegend = TRUE) %>% 
  layout(title = "Life Expectancy by Year",
         yaxis = list(title = "life expectancy"))

p1

```

```{r}
#| echo: false
#| eval: false

p1 <- mydata %>%
  plot_ly(x = ~year, 
          y = ~life_expectancy, 
          color = ~region,
          size = ~population, 
          text = ~country, 
          hoverinfo = "text")

p1 %>% 
  # add_markers(color = I("grey"), 
  #             alpha = 0.1,
  #             name = "All Years",
  #             type = 'scatter',
  #             mode = 'lines') %>%
  add_markers(frame = ~year, 
              ids = ~country) %>%
  animation_opts(1000, 
                 redraw = FALSE) %>% 
  layout(title = "Life Expectancy by Year",
         yaxis = list(title = "life expectancy"))



```

:::

## Life Expectancy by GDP

```{r}
#| warning: false
#| echo: true

p2 <- mydata %>%
  # filter(!is.na(GDP)) %>% 
  # filter(is.finite(GDP)) %>%
  plot_ly(x = ~GDP, 
          y = ~life_expectancy, 
          size = ~population, 
          fill = ~'',
          color = ~region,
          frame = ~year, 
          text = ~country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers',
          showlegend = TRUE) %>%
  layout(title = "Life Expectancy by GDP", 
         yaxis = list(title = "life expectancy"))


p2

```

## Life Expectancy by Logged GDP

Using logged GDP on the *x axis* means that we are looking at *relative*, instead of *absolute* changes in GDP.

::: {.panel-tabset}

### `ggplot` with `ggplotly`

```{r}
#| echo: true

p3 <- ggplot(mydata,
       aes(x = GDP,
           y = life_expectancy,
           color = region,
           size = population,
           frame = year,
           label = country)) + 
  geom_point() +
  labs(title = "Life Expectancy by GDP",
       x = "GDP",
       y = "Life Expectancy") +
  scale_color_discrete(name = "Region") +
  scale_x_log10()

ggplotly(p3)

```

### `plotly`

```{r}
#| warning: false
#| echo: true

p2 %>% 
  layout(xaxis = list(type = "log"))

```

:::


