---
title: "Research"
---

```{r}

library(readr) # get data

library(haven) # Stata data

library(ggplot2) # beautiful graphs

library(dplyr) # data wrangling

library(tidyr) # tidy data

library(DT) # nice data table

library(tibble) # tibbles

library(countrycode) # manage country data

library(plotly) # interactive graphs

library(maps) # maps

library(leaflet) # web maps

library(sp) # spatial data

library(sf) # simple features

```

## Research on Parenting and Child Development

Many of these publications are available from [Deep Blue](https://deepblue.lib.umich.edu/browse?type=author&value=Grogan-Kaylor%2C+Andrew). 

### Multiple Indicator Cluster Surveys (MICS) (UNICEF Data)

> This research is conducted in collaboration with colleagues [Shawna Lee](https://www.parentingincontext.org/), [Julie Ma](https://poverty.umich.edu/research/faculty-experts/julie-ma/), [Kaitlin Ward](https://www.kaitlinpward.com/), [Garrett Pace](https://www.garrettpace.com/), and [Jorge Cuartas](https://scholar.harvard.edu/jorgecuartas/home).

```{r}

global_data <- read_sf("../mapping/shapefiles/wrld_simpl/wrld_simpl.shp") 

# get MICS countries

country <- c("Afghanistan", "Algeria",  "Argentina",  
             "Bangladesh",  "Barbados",  "Belarus",  
             "Belize",  "Benin",  "Bosnia and Herzegovina",  
             "Cameroon",  "Central African Republic",  "Chad",  
             "Democratic Republic of the Congo",  
             "Republic of the Congo",  
             "Costa Rica",  "Cote d'Ivoire",  
             "Dominican Republic",  
             "El Salvador",  "Eswatini",  "Ghana",  
             "Guinea",  "Guinea Bissau",  "Guyana",  
             "Iraq",  "Jamaica",  "Kazakhstan",  
             "Kenya",  "Kosovo",  "Kyrgyzstan",  
             "Laos", "Macedonia",  "Madagascar",  
             "Malawi",  "Mali",  "Mauritania",  
             "Mexico",  "Moldova",  "Mongolia",  
             "Montenegro",  "Nepal",  "Nigeria",  
             "Pakistan",  "Panama",  "Paraguay",  
             "Sao Tome and Principe",  "Senegal",  "Serbia",  
             "Sierra Leone",  "Somalia",  "St. Lucia",  
             "State of Palestine",  "Suriname",  "Thailand",  
             "The Gambia",  "Togo",  "Trinidad and Tobago",  
             "Tunisia",  "Turkmenistan",  "Ukraine",  
             "Uruguay",  "Vietnam",  "Zimbabwe")

# convert to ISO3

country_iso <- countrycode(country, 
                           'country.name', 
                           'iso3c')

# MICS is an sf object that is subset of global_data

MICS <- global_data %>% 
  filter(ISO3 %in% country_iso)
  
```

```{r, fig.cap="Locations of Countries in MICS", fig.height=3}

library(leaflet)

leaflet() %>%
  setView(0, 25, zoom = 1.5) %>%
  # addTiles() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addPolygons(data = MICS, 
              fillOpacity = .75, 
              color = "#5b92e5", 
              label = MICS$NAME,
              highlightOptions = highlightOptions(color = "red", 
                                                  weight = 2,
                                                  bringToFront = TRUE)) 

```

#### Global Families Project

* [Project website](https://globalfamilies.quarto.pub/global-families-project/)

#### Publications

* "Associations between 11 parental discipline behaviours and child outcomes across 60 countries"
    + [Article](https://bmjopen.bmj.com/content/13/10/e058439)
    + [ArcGIS Storymap](https://arcg.is/1OSzvG0)
    + [HTML Slides](https://websites.umich.edu/~parentingandchilddev/MICS-discipline-11-forms/v1/MICS-discipline-11-forms.html)
    + [HTML Page](https://websites.umich.edu/~parentingandchilddev/MICS-discipline-11-forms/v2/MICS-discipline-11-forms.html)

* "Gender Inequality in Low- and Middle-Income Countries: Associations
with Parental Physical Abuse and Moderation by Child Gender"
    + [Article](https://www.mdpi.com/1841422)
    + [ArcGIS Storymap](https://arcg.is/1WzHHP)
    + [Press Release](https://news.umich.edu/gender-inequality-can-predict-high-rates-of-child-physical-abuse/)

* "The Association between Spanking and Physical Abuse of Young Children in 56 Low- and Middle-Income Countries"
    + [Article](https://doi.org/10.1016/j.chiabu.2022.105662)
    + [Press Release](https://news.umich.edu/physical-abuse-less-likely-when-spanking-is-eliminated/)
    + [HTML Slides](https://websites.umich.edu/~parentingandchilddev/MICS-physical-abuse/MICS-physical-abuse.html)
    + [ArcGIS Storymap](https://arcg.is/fq49)
    
* "Patterns of Caregiver Aggressive & Nonaggressive Discipline Toward Young Children in Low- and Middle-Income Countries: A Latent Class Approach"
    + [Article](https://doi.org/10.1016/j.chiabu.2022.105606)
    + [ArcGIS Storymap](https://arcg.is/1S5XPK)
    
* "Multilevel ecological analysis of the predictors of spanking across 65 countries"
    + [Article](https://bmjopen.bmj.com/content/11/8/e046075)
    + [ArcGIS Storymap](https://arcg.is/14CPKu)

* "Global Perspectives on Positive and Negative Parenting: A Bayesian Multilevel Analysis"
    + [Abstract](https://agrogan1.github.io/research/Bayes3MICS/)
    + [Article](https://doi.org/10.1177%2F0165025420981642)

* "Spanking and young children's socioemotional development in low- and middle-income countries"
    + [Open Access](https://doi.org/10.1016/j.chiabu.2018.11.003)
    + *Child Trends News Service* has released a well done [video summarizing this research](https://youtu.be/8LYZ-UpR5wE). 

### Adverse Childhood Experiences (ACES)

* ["Adverse childhood experiences and spanking have similar associations with early behavior problems"](https://doi.org/10.1016/j.jpeds.2021.01.072)

* ["Unpacking the impact of adverse childhood experiences on adult mental health"](http://dx.doi.org/10.1016/j.chiabu.2017.03.016) 

* ["Spanking and adult mental health impairment: The case for the designation of spanking as an adverse childhood experience"](https://doi.org/10.1016/j.chiabu.2017.01.014)  

### COVID-19

* ["Anxiety and Depression During COVID-19: Are Adults in Households with Children Faring Worse?"](https://link.springer.com/article/10.1007/s11606-021-07256-9)
    + [News Release](https://news.umich.edu/having-kids-at-home-may-reduce-pandemic-depression/)

### Colombia

* ["Physical Punishment as a Predictor of Early Cognitive Development: Evidence From Econometric Approaches"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7983059/)
* ["Physical punishment and Colombian children and adolescents’ cognitive and behavioral outcomes"](https://doi.org/https://doi.org/10.1016/j.appdev.2020.101140)
* ["Civil conflict, domestic violence, and poverty as predictors of corporal punishment in Colombia"](https://pubmed.ncbi.nlm.nih.gov/30772750/)

### Meta-Analysis of Research on Corporal Punishment

* [Meta-Analysis](https://agrogan1.github.io/research/cp/CP-meta.html) of Research on Corporal Punishment [[PubMed Full Text](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7992110/)] [[Slides](https://agrogan1.github.io/research/cpmeta-slides/cpmeta-slides.html)]

### Corporal Punishment Bans

* Data Visualizations of [Country Level Bans on Corporal Punishment](https://agrogan1.github.io/research/cpbans/)

```{r, fig.cap="Country Level Bans on Corporal Punishment"}

mydata <- read_csv("../research/cpbans/CPBans.csv") # read data

plot_geo(mydata) %>% 
  add_trace(locations = ~country_code, 
            color = ~year.of.prohibition,
            z = ~year.of.prohibition, 
            text = ~paste(country, 
                          "<br>year of prohibition:", 
                          year.of.prohibition),
            marker = list(size = 10)) %>%
  layout(geo = list(showland = FALSE,
                    showcountries = TRUE,
                    projection = list(type = 'orthographic',
                                      rotation = list(lon = -30,
                                                      lat = 10,
                                                      roll = 0)
                                      ))) %>%
  colorbar(title = 'year of prohibition') 

```


### Earlier Research on Corporal Punishment

* [Earlier Research On Corporal Punishment](https://agrogan1.github.io/research/cp/)

## Community Based Research in Alaska

* Culturally Grounded Strategies for Suicide and Alcohol Risk Prevention Delivered by Rural Alaska Native Communities: A Dynamic Wait‐listed Design Evaluation of the Qungasvik Intervention. 

    + [Open Access Article](https://onlinelibrary.wiley.com/doi/full/10.1002/ajcp.12621)
    
    
    
    
    