---
title: "Making a 'Big Number' (And Circle)"
subtitle: "With R"
author: "Andy Grogan-Kaylor"
date: "2026-3-23"
categories: [R, stats, dataviz]
---

# Background

A big number, enclosed by a circular bar graph, can be a compelling data visualization.

This post provides code for creating such data visualizations. The use of code initially makes this work more difficult than if using a drag-and-drop interface, but ultimately automates the work and makes it auditable and replicable.

# Example Using 30% and Michigan Blue

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

library(ggplot2) # beautiful graphs

ggplot(data = NULL, # NULL data set
       aes(x = 3, # arbitrary x; 3 is good
           y = 30)) + # y value to be graphed
  geom_col(fill = "#00274C") + # column & column color
  coord_polar(theta = "y") + # polar coordinates
  xlim(0, 4) + # xlimits create a doughnut
  ylim(0, 100) + # ylimits govern the circumference
  geom_text(aes(x = 0, # text in center
                y = 0,
                label = "30%"), # label for text
            size = 30, # text size
            col = "#00274C") + # text color
  theme_void() # empty theme

```

# Example Using 81% And Michigan Maize

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

library(ggplot2) # beautiful graphs

ggplot(data = NULL, # NULL data set
       aes(x = 3, # arbitrary x; 3 is good
           y = 81)) + # y value to be graphed
  geom_col(fill = "#ffcb05") + # column & column color
  coord_polar(theta = "y") + # polar coordinates
  xlim(0, 4) + # xlimits create a doughnut
  ylim(0, 100) + # ylimits govern the circumference
  geom_text(aes(x = 0, # text in center
                y = 0,
                label = "81%"), # label for text
            size = 30, # text size
            col = "#ffcb05") + # text color
  theme_void() # empty theme

```


