---
title: "Simpson's Paradox and The Aggregation Paradox"
subtitle: "Closely Related Statistical Phenomena"
author: "Andy Grogan-Kaylor" 
date: "2025-6-1"
categories: [stats, multilevel, science, teaching]
---

I am starting to think that [Simpson's Paradox](https://agrogan1.github.io/posts/Simpsons-paradox-simple-visual-explanation/) [@Simpson1951] and the Aggregation Paradox [C.F. @Gelman2007; @Nieuwenhuis2015] (which elsewhere I term "multilevel structure"), are variations of the same phenomenon.

Consider a simple regression equation with independent variable $x$ and a set of $k$ indicator variables $z$.

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + \sum_2^k \beta_k z_{kij} + u_{0j} + e_{ij}$$

* *Simpson's Paradox* could possibly occur if we omit the $z_{kij}$ from the regression equation.
* The *Aggregation Paradox* could possibly occur if we omit $u_{0j}$ from the regression equation.


```{r}
#| output: false
#| warning: false

set.seed(1080)

library(ggplot2)

library(ggthemes)

# library(gganimate)

library(ggsci)

library(patchwork)

```

```{r}
#| label: mydata

intervention <- seq(1,15) # intervention

group <- rep(seq(1,5), each=3) # group

index <- rep(seq(1,3), 5) # index within group

e <- rnorm(15, 0, .25) # randomly distributed error

outcome <- index + -.1 * intervention # + e

group <- factor(group)

mydata <- data.frame(intervention, group, outcome)

```

```{r}

p0 <- ggplot(mydata,
             aes(x = intervention,
                 y = outcome)) + 
  geom_point(aes(color=group), 
             size = 5) +
  geom_smooth(aes(color=group), 
              method="lm",
              linewidth = 2,
              se = FALSE) +
  geom_smooth(method="lm",
              linewidth = 2,
              se = FALSE) +
  # scale_color_viridis_d(name = "group",
  #                       option = "turbo") +
  # scale_color_bmj() +
  # scale_color_npg() +
  theme_minimal()

```

```{r}
#| label: fig-p0
#| message: false
#| fig-height: 7

name2 <- paste("school, \nneighborhood, or \ncountry expression(u_0)",
               expression(u_0))

(p0 + 
    labs(title = "Simpson's Paradox") +
   scale_color_npg(name = "group")) /
  (p0 + 
     labs(title = "Aggregation Paradox") +
     scale_color_npg(name = "school \n neighborhood \ncountry") +
   labs(caption = "A true multilevel model would have many more groups")) +
  plot_annotation(title = "A Model That Ignores Group, School, Neighborhood or Country Shows A Downward Slope",
                  subtitle = "A Model Accounting For Group, School, Neighborhood or Country Shows An Upward Slope",
                  theme = theme(title = element_text(size = rel(.75)))
) 

```