---
title: "An Updated Network Graph"
author: "Andy Grogan-Kaylor" 
date: "2026-7-5"
categories: [stats, dataviz, R, exploring data]
echo: true
code-fold: true
code-summary: "Show the code"
highlight-style: haddock
---

In a [previous post](../what-is-related-to-what/), I looked at some ways to get an idea of what *scales* and *demographic* variables in my data are related to other *scales* and *demographic* variables in my data. 

Among several methods, I looked at a *network graph* using `igraph` [@Csardi2006; @Csardi2025]. However, the network graph in my previous post was less than satisfactory because of the limited number of scales in my data. 

Below, I provide an example of a network graph with data with more scales.

# Call Libraries

```{r}

library(igraph) # network analysis

library(ggplot2) # beautiful graphs

library(ggraph) # network graphs for ggplot

library(faux) # simulated data

library(pander) # nice tables

library(skimr) # descriptive statistics

```

# Simulated Data

I simulate some data. 

```{r}

set.seed(54321) # random seed

# simulate data

cmat <- c(1.00, 0.30, 0.25, 0.15, 0.15, 0.15, # correlation matrix
          0.30, 1.00, 0.00, 0.50, 0.10, 0.15,
          0.25, 0.00, 1.00, 0.50, 0.50, 0.15,
          0.15, 0.50, 0.50, 1.00, 0.00, 0.15,
          0.15, 0.10, 0.50, 0.00, 1.00, 0.15,
          0.15, 0.15, 0.15, 0.15, 0.15, 1.00)

mydata <- rnorm_multi(n = 100, 
                   mu = c(0, 20, 100, 100, 100, 100),
                   sd = c(1, 5, 5, 5, 5, 5),
                   r = cmat, # correlation matrix
                   varnames = c("scaleA", 
                                "scaleB", 
                                "scaleC", 
                                "scaleD", 
                                "scaleE", 
                                "scaleF"),
                   empirical = FALSE)

```

::: {.callout-note collapse="true"}
# Look At The Data

```{r}

pander(head(mydata)) # nicely formatted table

```


```{r, skimr_include_summary = FALSE}

skim(mydata) # descriptive statistics

```
:::

# Network Graph

I provide code for labels, but because labels sometimes overlap and are thus difficult to read, I have commented out this code in the `igraph` code. I use a *Fruchterman-Reingold* layout [@ggraph; @Csardi2006; @Csardi2025] in each case to try to highlight the connections and centrality of different graph nodes. 

::: {.panel-tabset}

## `igraph`

```{r}
#| fig-height: 6
#| fig-width: 8

cor_mat <- cor(mydata, use = "complete.obs")

diag(cor_mat) <- 0 # set diagonal to 0

g <- graph_from_adjacency_matrix(cor_mat, # make network graph
                                 mode = "undirected", 
                                 weighted = TRUE)

# coords <- layout_in_circle(g) # get circular layout

coords <- layout.fruchterman.reingold(g) # Fruchterman-Reingold layout

par(mar=c(0, 0, 1, 0) + 1.5) # plot margins

absweight <- abs(E(g)$weight) # absolute value of weight

plot(g, 
     margin = 0.5,
     edge.width = absweight * 20, # absolute value
     vertex.size = 15,
     vertex.label.dist = 3.5,
     vertex.shape = "circle",
     vertex.color = "navy",
     # edge.label.color = "red",
     # edge.label = round(E(g)$weight, 2),
     # edge.label.dist = 1, 
     layout = coords, # layout
     edge.curved = 0,
     main = "Relationship of Scales in Our Data")

```

## `ggraph`

```{r}

ggraph(g, layout = 'fr') + 
  geom_edge_link(aes(width = .25 * weight,
                    color = weight,
                    label = round(weight, 2)),
                label_colour = "red",
                strength = .1,
                angle_calc = 'along',
                show.legend = FALSE) +
  geom_node_point(size = 7,
                  color = "navy") +
  geom_node_text(aes(label = name),
                  nudge_y = .15,
                  fill = "white") +
  scale_edge_color_gradient(name = "weight",
                            low = "lightgrey",
                            high = "darkgrey") +
  labs(title = "Relationship of Scales in Our Data") +
  theme_void() 

```

:::





