---
title: "An Updated Network Graph"
author: "Andy Grogan-Kaylor" 
date: "2026-6-10"
categories: [stats, dataviz, R]
echo: true
code-fold: true
code-summary: "Show the code"
highlight-style: haddock
---

In a [previous post](../what-is-related-to-what/), I looked at some ways to get an idea of what *scales* and *demographic* variables in my data are related to other *scales* and *demographic* variables in my data. 

Among several methods, I looked at a *network graph* using `igraph` [@Csardi2006; @Csardi2025]. However, the network graph in my previous post was less than satisfactory because of the limited number of scales in my data. 

Below, I provide an example of a network graph with data with more scales.

# Call Libraries

```{r}

library(igraph) # network analysis

library(ggplot2) # beautiful graphs

library(ggraph) # network graphs for ggplot

library(faux) # simulated data

library(pander) # nice tables

library(skimr) # descriptive statistics

```

# Simulated Data

I simulate some data. 

```{r}

set.seed(54321) # random seed

# simulate data

cmat <- c(1.00, 0.30, 0.25, 0.15, 0.15, 0.15,
          0.30, 1.00, 0.00, 0.50, 0.10, 0.15,
          0.25, 0.00, 1.00, 0.50, 0.50, 0.15,
          0.15, 0.50, 0.50, 1.00, 0.00, 0.15,
          0.15, 0.10, 0.50, 0.00, 1.00, 0.15,
          0.15, 0.15, 0.15, 0.15, 0.15, 1.00)

mydata <- rnorm_multi(n = 100, 
                   mu = c(0, 20, 100, 100, 100, 100),
                   sd = c(1, 5, 5, 5, 5, 5),
                   r = cmat, # correlation matrix
                   varnames = c("scaleA", 
                                "scaleB", 
                                "scaleC", 
                                "scaleD", 
                                "scaleE", 
                                "scaleF"),
                   empirical = FALSE)

```

::: {.callout-note collapse="true"}
# Look At The Data

```{r}

pander(head(mydata)) # nicely formatted table

```


```{r, skimr_include_summary = FALSE}

skim(mydata) # descriptive statistics

```
:::

# Network Graph

::: {.panel-tabset}

## `igraph`

```{r}
#| fig-height: 6
#| fig-width: 8

cor_mat <- cor(mydata, use = "complete.obs")

diag(cor_mat) <- 0 # set diagonal to 0

g <- graph_from_adjacency_matrix(cor_mat, # make network graph
                                 mode = "undirected", 
                                 weighted = TRUE)

coords <- layout_in_circle(g) # get circular layout

par(mar=c(0, 0, 1, 0) + 1.5) # plot margins

absweight <- abs(E(g)$weight) # absolute value of weight

plot(g, 
     margin = 0.5,
     edge.width = absweight * 20, # absolute value
     vertex.size = 15,
     vertex.label.dist = 2.5,
     vertex.shape = "circle",
     vertex.color = "navy",
     edge.label.color = "red",
     edge.label = round(E(g)$weight, 2),
     edge.label.dist = 1, 
     layout = coords, # layout
     edge.curved = 0,
     main = "Correlation of Scales in Our Data")

```

## `ggraph`

```{r}

ggraph(g, layout = 'linear', circular = TRUE) + 
  geom_edge_arc(aes(width = .25 * weight,
                    color = weight,
                    label = round(weight, 2)),
                label_colour = "red",
                strength = .1,
                angle_calc = 'along') +
  geom_node_point() +
  geom_node_label(aes(label = name),
                  nudge_y = .025,
                  fill = "white") +
  scale_edge_color_gradient(name = "weight",
                            low = "lightblue",
                            high = "darkblue") +
  labs(title = "Network Graph of Reports") +
  theme_void() 

```

:::





