---
title: "A Chord Diagram With `ggplot`"
subtitle: "Using The New `ggchord2` Library"
author: "Andy Grogan-Kaylor" 
date: "2026-8-25"
categories: [stats, dataviz, R, exploring data]
echo: true
code-fold: true
code-summary: "Show the code"
highlight-style: haddock
draft: false
---

In [other earlier posts](https://agrogan1.github.io/blog.html#category=exploring%20data), I have looked at some ways to get an idea of what *scales* and *demographic* variables in my data are related to other *scales* and *demographic* variables in my data. One way of visualizing these relationships is a [chord diagram](https://agrogan1.github.io/posts/chord-diagram/).

In this post, I try creating a *chord diagram* using the fantastic new `ggchord2` [@rennie2026; @ggchord2].

# Call Libraries

```{r}

library(faux) # simulated data

library(pander) # nice tables

library(skimr) # descriptive statistics

library(circlize) # circular graphs and chord diagrams

library(scales) # viridis color palettes

library(ggplot2) # beautiful graphs

library(ggchord2) # chord diagrams

```

# Simulated Data

I again simulate some data. 

```{r}

set.seed(54321) # random seed

# simulate data

cmat <- c(1.00, 0.30, 0.25, 0.15, 0.15, 0.15,
          0.30, 1.00, 0.00, 0.50, 0.10, 0.15,
          0.25, 0.00, 1.00, 0.50, 0.50, 0.15,
          0.15, 0.50, 0.50, 1.00, 0.00, 0.15,
          0.15, 0.10, 0.50, 0.00, 1.00, 0.15,
          0.15, 0.15, 0.15, 0.15, 0.15, 1.00)

mydata <- rnorm_multi(n = 100, 
                   mu = c(0, 20, 100, 100, 100, 100),
                   sd = c(1, 5, 5, 5, 5, 5),
                   r = cmat, # correlation matrix
                   varnames = c("scaleA", 
                                "scaleB", 
                                "scaleC", 
                                "scaleD", 
                                "scaleE", 
                                "scaleF"),
                   empirical = FALSE)

```

::: {.callout-note collapse="true"}
# Look At The Data

```{r}

pander(head(mydata)) # nicely formatted table

```


```{r, skimr_include_summary = FALSE}

skim(mydata) # descriptive statistics

```
:::

# Chord Diagram

For `ggchord` to work, I need data on the *correlations* of different scales, but need this correlation data to be formatted as a list of *sources* and *destinations.* 

I therefore first create a correlation matrix of my data, and then reformat this correlation matrix as a data frame of *sources*, *destinations*, and *correlations.*

[https://www.tutorialspoint.com/](https://www.tutorialspoint.com/article/how-to-convert-the-correlation-matrix-into-a-data-frame-with-combination-of-variables-and-their-correlations-in-r) has a page suggesting how to achieve this using *Base R*.

```{r}

cor_mat <- cor(mydata, # correlation matrix
               use = "complete.obs")

diag(cor_mat) <- 0 # set diagonal to 0

# thanks tutorialspoint.com!
mychorddata <- as.data.frame(as.table(cor_mat))

head(mychorddata) # look at the data

```

Using `ggchord` is then relatively straightforward. I use `geom_chord_diagram` which adds all of the required geometries of a chord diagram, but could use other `geoms` to add each geometry separately.

```{r}
#| fig-height: 6
#| fig-width: 8
#| label: fig-chord-diagram
#| fig-cap: "a chord diagram "

# pal_viridis()(5) # 5 viridis colors

ggplot(data = mychorddata, 
       mapping = aes(
         source = Var1,
         target = Var2,
         freq = Freq,
         fill = Var1)) + 
  geom_chord_diagram() + # add all the chord geoms
  # geom_chord_arcs() + # selecting specific 
  # geom_chord_sectors() + # chord geoms allows 
  # geom_chord_labels() + # for more customization
  scale_fill_viridis_d(name = "scale") + # nice colors
  labs(title = "Relationship of Scales In Our Data") +
  theme_void() +
  coord_fixed(clip = "off") # fixed coordinates; don't clip labels

```









