---
title: "A Chord Diagram"
author: "Andy Grogan-Kaylor" 
date: "2026-6-14"
categories: [stats, dataviz, R]
echo: true
code-fold: true
code-summary: "Show the code"
highlight-style: haddock
draft: true
---

In a [previous](../what-is-related-to-what/) [posts](../network-graph), I looked at some ways to get an idea of what *scales* and *demographic* variables in my data are related to other *scales* and *demographic* variables in my data. 

In this post, I try using a *chord diagram* using `chordDiagram` from the `circlize` library.

# Call Libraries

```{r}

library(faux) # simulated data

library(pander) # nice tables

library(skimr) # descriptive statistics

library(circlize) # circular graphs and chord diagrams

```

# Simulated Data

I simulate some data. 

```{r}

set.seed(54321) # random seed

# simulate data

cmat <- c(1.00, 0.30, 0.25, 0.15, 0.15, 0.15,
          0.30, 1.00, 0.00, 0.50, 0.10, 0.15,
          0.25, 0.00, 1.00, 0.50, 0.50, 0.15,
          0.15, 0.50, 0.50, 1.00, 0.00, 0.15,
          0.15, 0.10, 0.50, 0.00, 1.00, 0.15,
          0.15, 0.15, 0.15, 0.15, 0.15, 1.00)

mydata <- rnorm_multi(n = 100, 
                   mu = c(0, 20, 100, 100, 100, 100),
                   sd = c(1, 5, 5, 5, 5, 5),
                   r = cmat, # correlation matrix
                   varnames = c("scaleA", 
                                "scaleB", 
                                "scaleC", 
                                "scaleD", 
                                "scaleE", 
                                "scaleF"),
                   empirical = FALSE)

```

::: {.callout-note collapse="true"}
# Look At The Data

```{r}

pander(head(mydata)) # nicely formatted table

```


```{r, skimr_include_summary = FALSE}

skim(mydata) # descriptive statistics

```
:::

# Chord Diagram

```{r}
#| fig-height: 6
#| fig-width: 8

grid.col = c(scaleA = "red", 
             scaleB = "green", 
             scaleC = "blue", 
             scaleD = "orange", 
             scaleE = "purple")

cor_mat <- cor(mydata, use = "complete.obs") # correlation matrix

diag(cor_mat) <- 0 # set diagonal to 0

chordDiagram(cor_mat, 
             annotationTrack = c("name", "grid"), 
             big.gap = 25,
             scale = TRUE, 
             grid.col = grid.col)

```









