---
title: "Simpson's Paradox"
author: "Andy Grogan-Kaylor"
date: "2024-7-9"
categories: [stats, science, R]
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(dplyr)

library(tidyr)

library(ggplot2)

library(ggmosaic)

library(michigancolors)

library(pander)

library(foreign)

```

# Introduction

*Simpson's Paradox* [@Simpson1951] is the idea that associations between variables that are found between two variables in the sample as a whole, can be *very different* (or even *reversed*) when a third variable is introduced.

There are specific mathematical conditions under which Simpson's Paradox applies. However, thinking through these mathematical conditions may not be helpful to develop intuitions about Simpson's Paradox.

Careful inspection of visualizations can help us to develop some understanding of Simpson's Paradox.

> The key substantive conclusion is that models of the data that only account for two variables may provide *very different*--or even *opposite*--conclusions about a treatment or intervention than models of the data that account for many variables.

# Some Hypothetical Data

These data are based on the hypothetical data provided by @Simpson1951.

```{r}

count <- c(4, 3, 8, 5, 2, 3, 12, 15)

treatment <- c("untreated", "untreated", "treated", "treated",
               "untreated", "untreated", "treated", "treated")

treatment <- factor(treatment, levels = c("untreated", "treated"))

status <- c("alive", "dead", "alive", "dead",
            "alive", "dead","alive", "dead")

status <- factor(status, levels = c("dead", "alive"))

group <- c("A", "A", "A", "A", "B", "B", "B", "B")

group <- factor(group)

mydata <- data.frame(count, treatment, status, group)

write.dta(mydata, file = "simpsonsparadox.dta")

pander(mydata)

```

```{r}

mydata_expanded <- mydata %>% uncount(weights = count)

```

```{r}

# mosaicplot(table(mydata_expanded$treatment, 
#                  mydata_expanded$status),
#            col = c("darkred", "darkgreen"))

```

# Sample As A Whole

> The treatment appears to have NO effect. An equal percentage of those receiving and not receiving treatment are alive.

## Cross Tabulation

```{r}

pander(table(mydata_expanded$treatment, 
             mydata_expanded$status))

```

## Mosaic Plot

> Mosaic Plots are a little bit counterintuive at first. However, I believe that--after a little bit of study--Mosaic Plots provide the best visual representation of these relationships.

```{r}

ggplot(mydata_expanded) + 
  geom_mosaic(aes(x = product(status, treatment), 
                  fill=status), 
              na.rm=TRUE) +
  labs(title = "The Treatment Appears To Have NO Effect",
       subtitle = "In The Sample As A Whole",
       caption = "An Equal % Of Those Receiving And Not Receiving Treatment Are Alive") +
  scale_fill_manual(values = michigancolors()) +
  theme_bw()

```

## Pie Chart

> A Pie Chart sometimes is more intuitive at first, but actually has less information than a Mosaic Chart.

```{r, eval=TRUE}

ggplot(mydata_expanded,
       aes(x = 1, 
           fill = status)) +
  geom_bar(position=position_fill()) +
  labs(title = "The Treatment Appears To Have NO Effect",
       subtitle = "In The Sample As A Whole",
       caption = "An Equal % Of Those Receiving And Not Receiving Treatment Are Alive") +
  coord_polar(theta = "y") +
  facet_wrap(~treatment) +
  scale_fill_manual(values = michigancolors()) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```

# Sample Divided By Groups

> The treatment appears to have an effect. A greater percentage of those recieving treatment are alive.

## Cross Tabulation

```{r}

pander(table(mydata_expanded$treatment, 
             mydata_expanded$status, 
             mydata_expanded$group))

```

## Mosaic Plot

```{r}

ggplot(mydata_expanded) + 
  geom_mosaic(aes(x = product(status, treatment), 
                  fill=status), 
              na.rm=TRUE) +
  labs(title = "The Treatment DOES Appears To Have An Effect",
       subtitle = "When Examined By Group",
       caption = "A Greater % Of Those Recieving Treatment Are Alive") +
  facet_wrap(~group) +
  scale_fill_manual(values = michigancolors()) +
  theme_bw()

```

## Pie Chart

```{r, eval=TRUE}

ggplot(mydata_expanded,
       aes(x = 1, 
           fill = status)) +
  geom_bar(position=position_fill()) +
  labs(title = "The Treatment DOES Appears To Have An Effect",
       subtitle = "When Examined By Group",
       caption = "A Greater % Of Those Recieving Treatment Are Alive") +
  coord_polar(theta = "y") +
  facet_wrap(~group + treatment, nrow = 1) +
  theme_bw() +
  scale_fill_manual(values = michigancolors()) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```


