---
title: "What Is Related To What In My Data?"
author: "Andy Grogan-Kaylor" 
date: "2026-6-3"
categories: [stats, dataviz, R]
echo: true
code-fold: true
code-summary: "Show the code"
highlight-style: haddock
---

In some recent projects, I have had to get a quick idea of what *scales* and *demographic* variables in my data are related to other *scales* and *demographic* variables. 

It seems to me that one of the quickest ways to do this is with the `corrgram` [@corrgram] library. I also like `pairs.panels` in the `psych` [@psych] library, and `ggpairs` in the `GGally` [@GGally] library. 

Lastly, a *network graph* using `igraph` [@Csardi2006; @Csardi2025] requires substantially more programming, and produces a quite different looking visualization that may or may not be useful in some situations.

# Call Libraries

```{r}

library(MASS) # simulate multivariate data

library(pander) # nice tables

library(corrgram) # corrgrams

library(GGally) # pairs plots

library(psych) # useful routines for psych data

```


# Simulated Data

I simulate some data. 

::: {.callout-tip}
## Key Idea

Of particular note are that the strongest correlation is between `scaleA` and `scaleB`, and that `scaleA` and `scaleC` are negatively correlated.
:::

```{r}

set.seed(54321)

# simulate data

mu <- c(scaleA = 10, scaleB = 25, scaleC = 100) # means

Sigma <- matrix(c(1.0, 0.75, -0.3, # variance covariance
                  0.75, 1.0, 0.4,
                  -0.3, 0.4, 1.0), 
                nrow = 3, ncol = 3)

mydata <- mvrnorm(n = 500, 
                  mu = mu, 
                  Sigma = Sigma) # variance covariance

```

::: {.callout-note collapse="true"}
# Look At The Data

```{r}

pander(head(mydata)) # nicely formatted table

```


```{r, skimr_include_summary = FALSE}

skimr::skim(mydata) # descriptive statistics

```
:::

# What Is Related To What In My Data?

::: {.panel-tabset}

## `corrgram`

```{r}

corrgram(mydata, 
         order=TRUE, 
         lower.panel = panel.pie, # lower panel is pie charts
         upper.panel = panel.cor, # upper panel is correlations
         text.panel = panel.txt, # text panel is variable
         diag.panel=panel.minmax, # minimum and maximum
         main="Correlation of Scales in Our Data") 

```

## `ggpairs`

```{r}

# ggpairs(mydata) # simple default plot

ggpairs(mydata, # pairs plot
        lower = list(continuous = wrap("points", 
                                       color = "blue", 
                                       alpha = 0.5)),
        diag = list(continuous = wrap("densityDiag",
                                      color = "blue")),
        upper = list(continuous = wrap("cor",
                                       color = "black"))) + 
  labs(title = "Correlation of Scales in Our Data")

```

## `pairs.panels`

```{r}

pairs.panels(mydata, # pairs plot
             show.points = FALSE, # don't show data points
             main="Correlation of Scales in Our Data") 

```

## Network Graph

```{r}
#| fig-height: 6

library(igraph) # network analysis

cor_mat <- cor(mydata, use = "complete.obs")

diag(cor_mat) <- 0 # set diagonal to 0

g <- graph_from_adjacency_matrix(cor_mat, # make network graph
                                 mode = "undirected", 
                                 weighted = TRUE)

# par(mar=c(0, 0, 1, 0) + 1.5) # plot margins

absweight <- abs(E(g)$weight) # absolute value of weight

plot(g, 
     margin = 0.5,
     edge.width = absweight * 20, # absolute value
     vertex.size = 15,
     vertex.label.dist = 2.5,
     edge.label.color = "red",
     edge.label = round(E(g)$weight, 2),
     edge.label.dist = 1, 
     # layout = coords,
     edge.curved = .5,
     main = "Correlation of Scales in Our Data")

```

:::
