---
title: "Bar Charts in ggplot2"
author: "Andy Grogan-Kaylor"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    fig_height: 2
    highlight: haddock
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document: 
    fig_height: 3
    highlight: haddock
    number_sections: yes
    theme: default
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
mainfont: Tw Cen MT
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE, 
                      fig.margin = TRUE)

```

# Background

<style>
.list-group-item.active, 
.list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: black;
    background-color: grey;
    border-color: blue;
}
</style>

While `ggplot2`, and the ideas of an underlying *"grammar of graphics"*, make some kinds of graphing easier, `ggplot2` can make other types of graphing more difficult. 

One often tricky type of graph is the bar chart.  I have come to think that part of the difficulty with thinking about bar charts in `ggplot2` is that sometimes <span style="background-color: #FFFF00">**three very different types of bar charts look similar**</span>.

# Introduction

**Many** bar charts look **something like** the bar chart below.

```{r, echo=FALSE}

x <- round(rnorm(10, 10, 3), digits = 0)

library(ggplot2)

ggplot(data = NULL,
       aes(x = x)) + 
  geom_bar(fill = "blue") +
  ylab("what does y represent?") 

```

However, there are actually three slightly separate underlying *"grammars of graphics"* that might underlie a bar chart:

1. Bar charts where the height of the bars is the **number of observations** in each category.
2. Bar charts where the height of the bars is the **average value of the y variable** for that category.
3. Bar charts where the height of the bars is the **actual value of the y variable** for that **individual observation**.

Let's look at each situation in turn, since each situation demands a slightly different syntax.

# Our Data

```{r, echo=FALSE}

x <- c(1, 2, 2, 3, 3, 3)

y <- c(10, 5, 9, 8, 9, 10)

mydata <- data.frame(x,y)

library(pander)

pander(mydata)

```

# Bar charts where the <span style="background-color: #FFFF00">height of the bars is the **number of observations** in each category</mark>.

```{r}

library(ggplot2)

```

```{r}

ggplot(mydata, # the data that I am using
       aes(x = x)) + # 'aesthetic' only includes x
  geom_bar(fill = "blue") # using bars to graph

```

## Stacked Bar Chart

A simple change to the above aesthetic yields a stacked bar chart. Note that *fill* now becomes part of the *aesthetic* so that color fill differentiates the parts of the bar.

```{r}

ggplot(mydata, # the data that I am using
       aes(x = 1, # x is 1 
           fill = factor(x))) + # fill is x as a factor
  geom_bar() + # using bars to graph
  scale_fill_discrete(name = "x") # modify name of legend

```

> We then return to an unstacked bar chart to consider the syntax for adding labels.

## Add Labels

Adding labels requires adding an extra geom, `geom_text`. We have to add a new, non-intuitive aesthetic to `geom_text` to tell it where the labels are located, and that they represent the count of observations in each category.

```{r}

ggplot(mydata, # the data that I am using
       aes(x = x)) + # 'aesthetic' only includes x
  geom_bar(fill = "blue") + # using bars to graph
  geom_text(stat = "count", 
            aes(label = ..count.., # text of the label
                y = ..count.. + .5), # location of the label
            size = 5) # size of the label

```

# Bar charts where the <span style="background-color: #FFFF00">height of the bars is the **average value of the y variable** for that category</span>.

```{r}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = x, # 'aesthetic' includes x
           y = y)) + # and y
  stat_summary(fun.y = mean, # summarizing y 
               geom = "bar", # with bars
               fill = "blue") +
  labs(y = "average y")

```

## What About a Line Chart?

A line chart requires the addition of `group` to the aesthetic, and the use of a different geometry.

```{r}

ggplot(mydata, # the data that I am using
       aes(x = x,
           y = y,
           group = 1)) + # add 'group'
   stat_summary(geom = "line", # with line
                fun.y = mean,
                color = "blue")

```

# Bar charts where the <span style="background-color: #FFFF00">height of the bars is the **actual value of the y variable** for that category</span>.

Here is where things get complicated. We try something we've tried before, but now adding *y* to the aesthetic.

```{r, eval=FALSE}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = x, # 'aesthetic' includes x
           y = y)) + # and y
  geom_bar()

```

This generates the following error message:  

> `Error: stat_count() must not be used with a y aesthetic.`

The reason that we are getting this error is that, by default, `geom_bar()` is trying to **count up** the number of *x* values, and in **counting up** the number of *x* values, `geom_bar()` does not know what to do with the *y* value.

So we change this...

```{r}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = x, # 'aesthetic' includes x
           y = y)) + # and y
  geom_bar(stat = "identity", # use ACTUAL y for bar height
           fill = "blue") 

```

If we go back and look at our data, we remember that we have some duplicate *x*'s, 

```{r, echo = FALSE}

pander(mydata$x)

```

so some of the bars are actually **over-printing**.

We can see this if we make the bars semi-transparent, and outline the bars with a different color.

```{r}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = x,
           y = y)) + # 'aesthetic' only includes x
  geom_bar(stat="identity", # use ACTUAL y for bar height
           fill = "blue", # fill
           color = "red", # outline
           alpha = .5) # transparency

```

# Thinking Through The Issue

What is the solution?  We may want to go back and look at our data to ensure that if we are using the actual *y* value for the height of the bars that we do **not** have duplicate values of *x* in our data.  

**Or**, we may want to have the bars represent the **average value of y** rather than the **actual values of y**, as we did in one of the examples above.

# Of Course The Problem Wouldn't Come Up If We Had Different Data, Without Those Duplicate *x* Observations

```{r, echo=FALSE}

xREVISED <- c(1, 2, 3)

yREVISED <- c(10, 5, 8)

mydataREVISED <- data.frame(xREVISED, yREVISED)

library(pander)

pander(mydataREVISED)

```

## And Then A Bar Chart Is Easy

```{r}

library(ggplot2)

ggplot(mydataREVISED, # the data that I am using
       aes(x = xREVISED, # 'aesthetic' includes x
           y = yREVISED)) + # and y
  geom_bar(stat = "identity", # use ACTUAL y for bar height
           fill = "blue")

```

## A Line Chart Is Easy Too

```{r}

library(ggplot2)

ggplot(mydataREVISED, # the data that I am using
       aes(x = xREVISED, # 'aesthetic' includes x
           y = yREVISED)) + # and y
  geom_line(stat = "identity", # use ACTUAL y for bar height
           color = "blue")

```

# And Then There Are Pie Charts

In some ways it is confusing, and in some ways it is helpful, that according to the underlying *grammar of graphics* in `ggplot2`, pie charts can be seen as **bar charts that use polar coordinates**. With that in mind, we take some of our earlier code, and add `coord_polar()`.

In the code below, I need to ensure that `ggplot2` sees *x* as a **factor**, and to use *x* as a **fill**.  **Fill** is a crucial piece of information in a pie chart.

```{r}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = 1, 
           fill = factor(x))) + # x is now the fill
  geom_bar() + # using bars to graph 
  coord_polar(theta = "y") + # polar coordinates
  theme_void() # get rid of distracting numbers
  
```

> Unfortunately, pie charts are deprecated in some circles, so support for pie charts is not very strong in `ggplot`. It is certainly possible to create a pie chart in `ggplot`, but adding labels to a pie chart ends up being very very difficult.

```{r, eval=FALSE, echo=FALSE}

mybreaks <- table(mydata$x) # calculate breaks in data

# calculate percentages
# proportion table, rounded to 2 digits, X 100

mypercentages <- round(prop.table(table(mydata$x)),
                       digits = 2) * 100


mylabels <- paste0(mybreaks, 
                  ", ",
                  mypercentages, 
                  "%") # paste together into labels

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = 1, 
           fill = factor(x))) + # x is now the fill
  geom_bar() + # using bars to graph 
  coord_polar(theta = "y") + # polar coordinates
  theme_void() + # get rid of distracting numbers
  scale_fill_viridis_d(name = "x",
                       breaks = mybreaks,
                       labels = mylabels)

```

## Pie Chart With Better Colors

```{r}

library(ggplot2)

ggplot(mydata, # the data that I am using
       aes(x = 1, 
           fill = factor(x))) + 
  geom_bar() + # using bars to graph 
  coord_polar(theta = "y") + # polar coordinates 
  scale_fill_viridis_d(name = "x") + # beautiful colors; name for legend
  theme_void() # get rid of distracting numbers

```

# Bar Chart With Better Colors

Up until now, we have had a *minimalist* vision of bar charts, where every bar is the same color, because color would not add **additional information**, over and above the information contained in the position on the *x* axis.  However, for the sake of design, we may also choose to add some color to our bar charts. `ggthemes`, `ggthemr` and `viridis` are all ways of adding color to `ggplot` graphs.

```{r}

library(ggplot2)

library(viridis) # wonderful colors

ggplot(mydata, # the data that I am using
       aes(x = x, # x is on the x axis
           fill = factor(x))) + # x is also a factor for fill
  geom_bar() + # using bars to graph 
  scale_fill_viridis_d(name = "x") # beautiful colors; name for legend

```

# More Examples

More examples can be found [here](https://agrogan1.github.io/dataviz/how-to-choose-a-chart/how-to-choose-a-chart-v3.html)








