---
title: "Making a 'Big Number' (And Circle)"
subtitle: "With R"
author: "Andy Grogan-Kaylor"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r, include = F}

# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming

knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)

library(flipbookr)

library(tidyverse)

library(xaringanExtra)

library(emo) # emoji

xaringanExtra::use_tile_view()

# xaringanExtra::use_slide_tone()

xaringanExtra::use_share_again()

```

```{css, echo = FALSE}

@import url('https://fonts.googleapis.com/css2?family=Montserrat&display=swap');
h1, h2, h3 {
  font-family: 'Montserrat', sans-serif;
}
body { 
    font-family: 'Montserrat', sans-serif;
}
.author, .date {
  font-family: 'Montserrat', sans-serif;
}

.title-slide { 
  color: #ffcb05;
  background-color: #00274C; 
}
.title-slide h1 {
    color: #ffcb05;
}
pre {
  white-space: pre-wrap;
}
blockquote {
  background: #FFFFFF;
  color: #000;
  border-left: 5px solid #ffcb05;
  margin: 1.5em 10px;
  padding: 0.5em 10px;
}

```

# How To Navigate This Presentation

* Use the `r emo::ji("left_arrow")` `r emo::ji("right_arrow")` keys to move through the presentation.
* Press *o* for *panel overview*.

--

# Background

Code for creating "big numbers" enclosed with a "circle graph". Use of code initially makes this work more difficult than if using a drag-and-drop interface, but ultimately automates the work and makes it auditable and replicable.

---

```{r graph1, include = FALSE}

library(ggplot2) # beautiful graphs

ggplot(data = NULL, # NULL data set
       aes(x = 3, # arbitrary x; 3 is good
           y = 30)) + # y value
  geom_col(fill = "#00274C") + # column
  coord_polar(theta = "y") + # polar coordinates
  xlim(0, 4) + # xlimits create a doughnut
  ylim(0, 100) + # ylimits govern the circumference
  geom_text(aes(x = 0, # text in center
                y = 0,
                label = "30%"), # label for text
            size = 30,
            col = "#00274C") +
  theme_void() # empty theme

```

---

```{r, code = knitr::knit_code$get("graph1"), eval = FALSE, echo = FALSE}
```

`r chunk_reveal("graph1", break_type = "auto", widths = c(4,2))`

---

```{r graph2, include = FALSE}

library(ggplot2) # beautiful graphs

ggplot(data = NULL, # NULL data set
       aes(x = 3, # arbitrary x; 3 is good
           y = 81)) + # y value
  geom_col(fill = "#ffcb05") + # column
  coord_polar(theta = "y") + # polar coordinates
  xlim(0, 4) + # xlimits create a doughnut
  ylim(0, 100) + # ylimits govern the circumference
  geom_text(aes(x = 0, # text in center
                y = 0,
                label = "81%"), # label for text
            size = 30,
            col = "#ffcb05") +
  theme_void() # empty theme

```

---

```{r, code = knitr::knit_code$get("graph2"), eval = FALSE, echo = FALSE}
```

`r chunk_reveal("graph2", break_type = "auto", widths = c(4,2))`

