---
title: "Raincloud Plots"
description: |
  An innovative format to show variable distributions.
author:
  - name: Andy Grogan-Kaylor 
    url: https://www.umich.edu/~agrogan
    affiliation: University of Michigan
    affiliation_url: https://www.umich.edu
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    highlight: haddock
    toc: yes
    code_folding: hide
bibliography: raincloud.bib
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

```

# Introduction

*Effectively* conveying the distribution of a variable is a difficult problem. There are many `geom`etries for showing distributions. However, showing the distribution of a variable in an intuitive fashion for decision makers is still difficult.

# Sample Data

```{r}

load("social_service_agency.RData")

```

# Call `ggplot`

```{r}

library(ggplot2)

```

# Some Common Distributions

## Histogram

```{r}

ggplot(clients,
       aes(x = mental_health_T2,
           fill = program)) + 
  geom_histogram() +
  scale_fill_viridis_d() +
  facet_wrap(~program) +
  labs(title = "Mental Health of Clients",
       subtitle = "by Program",
       x = "Mental Health At Time 2",
       caption = "Higher Scores Equal Better Mental Health") +
  theme_minimal()

```

## Density

```{r}

ggplot(clients,
       aes(x = mental_health_T2,
           fill = program)) + 
  geom_density() +
  scale_fill_viridis_d() +
  facet_wrap(~program) +
  labs(title = "Mental Health of Clients",
       subtitle = "by Program",
       x = "Mental Health At Time 2", 
       caption = "Higher Scores Equal Better Mental Health") +
  theme_minimal()

```

## Boxplot

```{r}

ggplot(clients,
       aes(x = mental_health_T2,
           fill = program)) + 
  geom_boxplot() +
  scale_fill_viridis_d() +
  facet_wrap(~program) +
  labs(title = "Mental Health of Clients",
       subtitle = "by Program",
       x = "Mental Health At Time 2",
       caption = "Higher Scores Equal Better Mental Health") +
  theme_minimal()

```
## Jittered Points

```{r}

ggplot(clients,
       aes(x = mental_health_T2,
           y = 1,
           color = program)) + 
  geom_jitter() +
  scale_color_viridis_d() +
  facet_wrap(~program) +
  labs(title = "Mental Health of Clients",
       subtitle = "by Program",
       x = "Mental Health At Time 2", 
       caption = "Higher Scores Equal Better Mental Health") +
  theme_minimal()

```
# Raincloud Plots

The Raincloud Plot [@raincloud] works by combining the last three of these geometries: a density; a boxplot, and jittered points. Below is one (relatively simplified) implementation of these ideas. More detailed implementations can be found in @raincloud.

```{r}

library(ggplot2)

library(ggdist)

ggplot(clients,
       aes(x = mental_health_T2,
           y = program)) +
  stat_slab(aes(fill = program),
            position = position_nudge(x = 0, 
                                      y = .20),
            scale = .5) +
  geom_boxplot(color = "grey",
               width = .20,
               size = 1,
               show.legend = FALSE) +
  geom_point(aes(color = program),
             position = position_jitter(height = .20),
             size = 1.0,
             alpha = .5) +
  labs(title = "Mental Health of Clients",
       subtitle = "by Program",
       x = "Mental Health At Time 2",
       caption = "Higher Scores Equal Better Mental Health") +
  scale_fill_viridis_d() + 
  scale_color_viridis_d() + 
  theme_minimal() 

```

