---
title: "Simpson's Paradox"
author: "Andy Grogan-Kaylor"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    number_sections: yes
    toc: yes
  distill::distill_article:
    highlight: haddock
    toc: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(dplyr)

library(tidyr)

library(ggplot2)

library(ggmosaic)

library(pander)

library(foreign)

```

# Introduction

*Simpson's Paradox* is the idea that associations between variables that are found between two variables in the sample as a whole, can be *very different* (or even *reversed*) when a third variable is introduced.

# Some Hypothetical Data

These data are based on the hypothetical data provided by Simpson (1951).

```{r}

count <- c(4, 3, 8, 5, 2, 3, 12, 15)

treatment <- c("untreated", "untreated", "treated", "treated",
               "untreated", "untreated", "treated", "treated")

treatment <- factor(treatment, levels = c("untreated", "treated"))

status <- c("alive", "dead", "alive", "dead",
            "alive", "dead","alive", "dead")

status <- factor(status, levels = c("dead", "alive"))

group <- c("A", "A", "A", "A", "B", "B", "B", "B")

group <- factor(group)

mydata <- data.frame(count, treatment, status, group)

write.dta(mydata, file = "simpsonsparadox.dta")

pander(mydata)

```

```{r}

mydata_expanded <- mydata %>% uncount(weights = count)

```

```{r}

# mosaicplot(table(mydata_expanded$treatment, 
#                  mydata_expanded$status),
#            col = c("darkred", "darkgreen"))

```

# Sample As A Whole

> The treatment appears to have no effect.

## Cross Tabulation

```{r}

pander(table(mydata_expanded$treatment, 
             mydata_expanded$status))

```

## Mosaic Plot

> Mosaic Plots are a little bit counterintuive at first. However, I believe they provide the best visual representation of these relationships.

```{r}

ggplot(mydata_expanded) + 
  geom_mosaic(aes(x = product(status, treatment), 
                  fill=status), 
              na.rm=TRUE) +
  labs(title = "The Treatment Appears To Have NO Effect",
       subtitle = "In The Sample As A Whole",
       caption = "An Equal % Of Those Receiving And Not Receiving Treatment Are Alive") +
  theme_bw()

```

## Pie Chart

```{r, eval=TRUE}

ggplot(mydata_expanded,
       aes(x = 1, 
           fill = status)) +
  geom_bar(position=position_fill()) +
  labs(title = "The Treatment Appears To Have NO Effect",
       subtitle = "In The Sample As A Whole",
       caption = "An Equal % Of Those Receiving And Not Receiving Treatment Are Alive") +
  coord_polar(theta = "y") +
  facet_wrap(~treatment) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```

# Sample Divided By Groups

> The treatment appears to have an effect.

## Cross Tabulation

```{r}

pander(table(mydata_expanded$treatment, 
             mydata_expanded$status, 
             mydata_expanded$group))

```

## Mosaic Plot

```{r}

ggplot(mydata_expanded) + 
  geom_mosaic(aes(x = product(status, treatment), 
                  fill=status), 
              na.rm=TRUE) +
  labs(title = "The Treatment DOES Appears To Have An Effect",
       subtitle = "When Examined By Group",
       caption = "A Greater % Of Those Recieving Treatment Are Alive") +
  facet_wrap(~group) +
  theme_bw()

```

## Pie Chart

```{r, eval=TRUE}

ggplot(mydata_expanded,
       aes(x = 1, 
           fill = status)) +
  geom_bar(position=position_fill()) +
  labs(title = "The Treatment DOES Appears To Have An Effect",
       subtitle = "When Examined By Group",
       caption = "A Greater % Of Those Recieving Treatment Are Alive") +
  coord_polar(theta = "y") +
  facet_wrap(~group + treatment, nrow = 1) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```

# Reference

Simpson, E. (1951). The Interpretation of Interaction in Contingency Tables. *Journal of the Royal Statistical Society. Series B (Methodological)*, 13(2), 238-241. Retrieved February 2, 2021, from [http://www.jstor.org/stable/2984065](http://www.jstor.org/stable/2984065)

