---
title: "Visualizing Categorical Data in R"
description: |
  Bar Graphs and Beyond
author:
  - name: Andy Grogan-Kaylor 
    url: https://agrogan1.github.io/
    affiliation: University of Michigan
    affiliation_url: https://www.umich.edu
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    highlight: haddock
    toc: yes
    code_folding: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(pander) # nice tables

```

# Introduction 

Visualizing categorical data presents unique challenges. A common solution is a bar graph, which may often be the best data visualization solution. 

However there are also some alternatives to bar graphs.

Below we present some options for bar graphs, and some possible alternative strategies.

# Some Data

```{r}

library(tibble) # rowise data frame (tibble) creation

library(tidyr) # data wrangling

mydata <- tribble(
  ~group, ~outcome, ~count,
  "Group A",   "beneficial outcome", 55,
  "Group A",   "undesirable outcome", 40,
  "Group B",   "beneficial outcome", 50,
  "Group B",   "undesirable outcome", 75
)

mydata$group <- factor(mydata$group) # data wrangling

mydata$outcome <- factor(mydata$outcome) # data wrangling

# duplicate the observations by count

mydata <- mydata %>% uncount(count) 

pander(table(mydata)) # nice table of data

```

# Call The Graphing Library

```{r}

library(ggplot2)

```

# Stacked Bar Graph

```{r}

ggplot(mydata, aes(x = group, # x is group
                   fill = outcome)) + # color fill is outcome
  geom_bar() + # bars
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_minimal() # nice theme

```

# Unstacked Bar Graph

```{r}

ggplot(mydata, aes(x = group, # x is group
                   fill = outcome)) + # color fill is outcome
  geom_bar(position = position_dodge()) + # "dodged" bars
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_minimal() # nice theme

```
# Faceted Bar Graph

```{r}

ggplot(mydata, aes(x = outcome, # x is outcome
                   fill = outcome)) + # color fill is outcome
  geom_bar() + # bars
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_minimal() + # nice theme
  facet_wrap(~group) # facet on group

```

# Pie Chart

```{r}

ggplot(mydata, aes(x = 1, # x is always 1
                   fill = outcome)) + # color fill is outcome
  geom_bar(position = "fill") + # bars
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_void() + # void theme for pie charts
  coord_polar(theta = "y") + # polar coordinates
  facet_wrap(~group) # facet on group

```

# Mosaic Plot

```{r}

library(ggmosaic) # mosaic plots

ggplot(mydata) + 
  geom_mosaic(aes(x = product(group), # "mosaic" geometry
                  fill = outcome)) +
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_minimal()  # nice theme

```


# Waffle Plot

```{r}

library(waffle) # waffle geometry

library(dplyr) # data wrangling

```

```{r}

# make a data set of counts

mycounts <- mydata %>%
  group_by(group, outcome) %>% # group by group & outcome
  tally() # count up observations

pander(mycounts) # replay this data

```

```{r}

# use geom_waffle with this data set of counts

ggplot(mycounts, # use this new data
       aes(fill = outcome, # color fill is outcome
           values = n)) + # values are n
  geom_waffle(color = "grey") + # waffle geometry w/ grey separator
  facet_wrap(~group) + # facet on group
  coord_equal() + # squares!
  scale_fill_viridis_d(option = "cividis") + # beautiful colors
  theme_void()  # nice theme

```







